---
title: "TMT Acetylome proteome"
authors: "Chandra Chilamakuri, Eva Papachristou, Sarah Harton"
date: "09/08/2022 "
output:
  html_document:
    code_folding: hide
    number_sections: no
    toc: yes
    fig_caption: yes
---

### Introduction



```{r load_lib, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(qPLEXanalyzer))
suppressPackageStartupMessages(library(sva))
suppressPackageStartupMessages(library(pander))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gplots))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(DT))

log2_plus_one_tnas <- function(x){
  
  return(log2(x + 1))
}

get_no_pep_and_proteins_identified <- function( intensities_data, run='run' ){
  
  int_tab <- Biobase::fData(intensities_data)
  no_of_pep <- nrow(int_tab)
  no_of_uniq_prot <- length(unique(int_tab$Accessions))
  higest_peptides <- as.data.frame(table(int_tab$Accessions) ) %>% 
    arrange( desc(Freq)) %>% 
    head(n=1)
  prote_with_highest_pep <- as.vector(higest_peptides$Var1)
  number_of_peptides <- higest_peptides$Freq
  median_peptides <- median(table(int_tab$Accessions) )
  
  tab <- data.frame(Run=run, 
                    Unique_peptides=no_of_pep,
                    Uniqe_proteins=no_of_uniq_prot,
                    Median_peptides_per_protein=median_peptides,
                    Protein_with_highest_peptides=prote_with_highest_pep,
                    Number_of_peptides=number_of_peptides
  )
  return(tab)
}

hmMissPlot <- function(MSnObj){
  exprs(MSnObj) %>%
    as_tibble() %>%
    mutate(PeptideID = str_c("Pep", row_number())) %>%
    nest(Data = where(is.numeric)) %>%
    mutate(Filter = Data) %>%
    unnest(Filter) %>%
    mutate(across(where(is.numeric), is.na)) %>%
    rename_with(.cols = where(is.logical), ~str_c("F", .x)) %>%
    mutate(MeanInt = map_dbl(Data, rowMeans, na.rm = TRUE)) %>%
    arrange(across(c(where(is.logical), MeanInt))) %>%
    select(Data) %>%
    unnest(Data) %>%
    rowid_to_column("Ord") %>%
    pivot_longer(names_to = "Sample", values_to = "Intensity", -Ord) %>%
    mutate(Intensity=log10(Intensity)) %>%
    ggplot(aes(x=Sample, y=Ord)) +
    geom_tile(aes(fill=Intensity)) +
    scale_fill_continuous(na.value = 'white') +
    labs(y=NULL, fill="log(Intensity)") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          axis.text.x = element_text(angle = 45, hjust = 1),
          panel.border = element_rect(colour = "black", fill=NA, size=1),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank()) +
    scale_x_discrete(expand = c(0,0)) +
    scale_y_discrete(expand = c(0,0))
}
```

```{r read_sample_metadata, message=FALSE, warning=FALSE}
# read sample sheet
s_sheet <- read.csv( file='../../../data/processed/sample_sheet_v1.0.csv', stringsAsFactors = F) 

datatable(s_sheet)

# read annotation
anno_tab <- read_tsv(file='../../../data/annotation_human_10_09_2021.txt', col_names = T) 

# read intensities
int_tab <- read.csv( file='../../../data/processed/reformatted_peptide_intensities_acetylome_09_08_2022_v1.0.csv', stringsAsFactors = F) %>% 
  #select( -Modifications) %>% 
  filter(Protein_Groups  == 1 ) %>% 
  filter(Proteins == 1 ) %>% 
  select(-Protein_Groups, -Proteins ) %>% 
  filter(!is.na(Accessions)) %>%  
  mutate( ID=str_c(Accessions,Sequence, Modifications, sep = '__')) %>% 
  filter(str_detect(ID, 'Acetyl') )  %>% 
  select(Sequence,Accessions, Modifications,ID, one_of(s_sheet$SampleName))  



MSnset_data1 <- convertToMSnset( ExpObj = int_tab, metadata =s_sheet, Sequences =1, 
                                 Accessions =2, indExpData = c(5:ncol(int_tab)), 
                                 rmMissing=FALSE )
```




### Before filtering: missing values plot

```{r missing_vales_plot, message=FALSE, warning=FALSE}
hmMissPlot(MSnset_data1)
```

### Before filtering: Peptide Stats 
Statistics table consisting of total peptides/proteins in each plex.

```{r pep_stats, message=FALSE, warning=FALSE}
stats_table <- get_no_pep_and_proteins_identified(intensities_data = MSnset_data1, run = '1' )
pander(stats_table,style="rmarkdown")
```



### After filtering: Missing valies

* currently filtered all the peptides with missing values 
* As we have knock-out geneotyle, needs to replace missing values with zeros in peptides with Acetyl modification? I am not sure. 

```{r af_missing_vales_plot, message=FALSE, warning=FALSE}
MSnset_data1 <- convertToMSnset( ExpObj = int_tab, metadata =s_sheet, Sequences =1, 
                                 Accessions =2, indExpData = c(5:ncol(int_tab)), 
                                 rmMissing=TRUE )
hmMissPlot(MSnset_data1)
```

### After filtering: Peptide Stats   

Statistics table consisting of total peptides/proteins in each plex.

```{r af_pep_stats, message=FALSE, warning=FALSE}
stats_table <- get_no_pep_and_proteins_identified(intensities_data = MSnset_data1, run = '1' )
pander(stats_table,style="rmarkdown")
```


### After filtering: Intensity plots

The plots below shows the peptide intensities before and after normalization. The data is normalized via median scaling. 

```{r af_intensityPlot1, warning=FALSE, message=FALSE, , fig.cap="Density plots of raw intensities for TMT run.", fig.height=4}
MSnset_data1_norm <- normalizeScaling(MSnset_data1, median)
p1 <- intensityPlot(MSnset_data1, title = "Before Normalisation", 
                    trFunc = log2_plus_one_tnas)
p2 <- intensityPlot(MSnset_data1_norm, title = "Normalisation", 
                    trFunc = log2_plus_one_tnas)
grid.arrange(p1, p2, ncol = 2, nrow = 1)
```

### After filtering:intensity Boxplots

```{r af_intensityBoxplot1, warning=FALSE, message=FALSE, , fig.cap="boxplots of raw intensities for TMT run.", fig.height=4}
p1 <- intensityBoxplot(MSnset_data1, title = "Before Normalisation")
p2 <- intensityBoxplot(MSnset_data1_norm, title = "After Normalisation")
grid.arrange(p1, p2, ncol = 2, nrow = 1)
```



### After filtering:Relative log intensity plot

```{r af_rliPlot1, warning=FALSE, message=FALSE, , fig.cap="Relative intensity plots of raw intensities for TMT run.", fig.height=4}
p1 <- rliPlot(MSnset_data1, title = "Before Normalisation")
p2 <- rliPlot(MSnset_data1_norm, title = "After Normalisation")
grid.arrange(p1, p2, ncol = 2, nrow = 1)

```



### After normalisation - reatin Acetyl peptides

- For normalization retain all the peptides, except peptides with missing values
- Then filter out all not Acetyl peptides


```{r message=FALSE, warning=FALSE, echo=FALSE}
# re-build the object after retaining only peptides with acetyl modifications
# read intensities
int_tab <- int_tab %>% 
  mutate(Accessions = ID)

MSnset_data1 <- convertToMSnset( ExpObj = int_tab, metadata =s_sheet, Sequences =1, 
                                 Accessions =2, indExpData = c(5:ncol(int_tab)), 
                                 rmMissing=TRUE )
MSnset_data1_norm <- normalizeScaling(MSnset_data1, median)

```

```{r intensity_plots, warning=FALSE, message=FALSE, , fig.cap="Density plots of normalized intensities for TMT run.", fig.height=4}

IDs <- fData(MSnset_data1_norm)$ID
acetyle_proteins <- str_remove(IDs, '__.*$') 
keep <- str_detect(IDs, 'Acetyl')
uniq_acetyle_pteins <- unique(acetyle_proteins)
min_ace <- range(table(acetyle_proteins) )[1]
max_ace <- range(table(acetyle_proteins) )[2]
median_ace <- median(range(table(acetyle_proteins) ))
MSnset_data1_Pnorm <- MSnset_data1_norm
intensityPlot(MSnset_data1_Pnorm)
```

* Number of peptides with acetyl modifications: `r nrow(MSnset_data1_norm)`
* Number of unique proteins with acetyl modifications: `r length(uniq_acetyle_pteins)`
* Range of acetyl modifications per protein: `r min_ace` - `r max_ace`
* Median acetyl modifications per protein: `r median_ace`

### PCA

```{r batchcorr2, warning=FALSE, message=FALSE, , fig.cap="PCA plots after  normalisation and summarisation", fig.height=12, fig.width=12, fig.asp=0.7}
comb_batch <- MSnset_data1_Pnorm
p1 <- pcaPlot(comb_batch, labelColumn="SampleName", pointsize=3, x.nudge=3,transform=TRUE,labelsize=3)
p1
p2 <- pcaPlot(comb_batch, labelColumn="date", pointsize=3, x.nudge=3,transform=TRUE,labelsize=3)
p2
```

### Sample correlation plot

The plot below shows the correlation between samples based on normalized protein intensities. 

```{r cor_plot_bn, warning=FALSE, message=FALSE, , fig.cap="Correlation plots for samples based on normalized protein intensities.", fig.height=8, fig.width=8}
corrPlot(comb_batch,addValues=FALSE)
```

### Hierarchical clustering analysis

The plot below shows hierarchical clustering plots for all samples using eucledian distance and complete linkage methods.

```{r hclust, warning=FALSE, message=FALSE, , fig.cap="Clusering plot of normalized protein intensities for TMT run.", fig.height=8, fig.width=8, fig.asp=0.5}
hierarchicalPlot(comb_batch)
```




### Differential Expression

A **limma** based statistical analysis is carried out to find differentially-expressed proteins. 

The plot below shows the differentially expressed protein (in color).

```{r diffexp1, warning=FALSE, message=FALSE, , fig.height=4, results='asis'}
contrasts <- c(
  compound_het_vs_WT = "compound_het - WT",
  Homo_knock_in_vs_WT = "Homo_knock_in - WT",
  Knockout_vs_WT = "Knockout - WT",
  Knockout_vs_Homo_knock_in = "Knockout - Homo_knock_in",
  Knockout_vs_compound_het = "Knockout - compound_het",
  compound_het_vs_Homo_knock_in = "compound_het - Homo_knock_in"

)

pData(comb_batch)$Run <- as.factor(pData(comb_batch)$Run)

exprs(comb_batch) <- log2(exprs(comb_batch) + 1)
diffstats <- computeDiffStats(comb_batch, contrasts=contrasts, batchEffect = c("date", "SampleGroup"), transform=FALSE)
diffexp <- list()
for (i in 1:length(contrasts)){
   de_tab <- getContrastResults(diffstats=diffstats, contrast=contrasts[i], writeFile= FALSE)
   
   de_tab <- de_tab %>% 
     mutate(Accessions = str_remove(Accessions, '__.*$')) %>% 
     dplyr::select(-ID)
   
   de_tab <- inner_join(anno_tab, de_tab, by='Accessions') %>% 
     arrange(adj.P.Val)
   diffexp[[i]] <- de_tab
   
   
   ## MA plot
   p_de_tab <- de_tab %>% 
     mutate(significant = adj.P.Val < 0.05)
   
   text_tab <- p_de_tab %>% 
     filter(adj.P.Val < 0.05)
   
   ma_p <- ggplot(data=p_de_tab, mapping = aes(x=AvgIntensity,y=log2FC, color=significant)) +
       geom_point(size=1, alpha=0.3) +
       geom_hline(yintercept = 0) +
     labs(
       title = contrasts[i]
     ) +
     theme_classic() +
     theme(
       plot.title = element_text( hjust = 0.5)
     )
   
   vol_p <- ggplot(data=p_de_tab, mapping = aes(x=log2FC,y=-log10(adj.P.Val), color=significant)) +
       geom_point(size=1, alpha=0.3) +
       geom_hline(yintercept = 0) +
     labs(
       title = contrasts[i]
     ) +
     theme_classic() +
     theme(
       plot.title = element_text( hjust = 0.5)
     )
   
   
   if(nrow(text_tab) == 0){
     print(ma_p)
     print(vol_p)
     
   } else {
     
     text_tab <- head(text_tab, n=10)
     ma_p <- ma_p +
       geom_point(data= text_tab, mapping = aes(x=AvgIntensity,y=log2FC), inherit.aes = F) +
       geom_text(data= text_tab, mapping = aes(x=AvgIntensity,y=log2FC, label=GeneSymbol), 
                 inherit.aes = F, size=3, color='blue', nudge_y=0.05)
     
     
     vol_p <- vol_p + 
       geom_point(data= text_tab, mapping = aes(x=log2FC,y= -log10(adj.P.Val)), inherit.aes = F) +
       geom_text(data= text_tab, mapping = aes(x=log2FC,y=-log10(adj.P.Val), label=GeneSymbol), 
                 inherit.aes = F, size=3, color='blue', nudge_y=0.05)
     
     
     print(ma_p)
     print(vol_p)
     
   }
   
   
}


```




## Differential Results 



Final statistics table consisting of total peptides/proteins and deferentially expressed proteins in each contrast (adj.P.Val < 0.05).


```{r table, warning=FALSE, message=FALSE}
stats_table <- matrix(0,1,length(contrasts)+1)
colnames(stats_table) <- c("Peptides",contrasts)
stats_table[,1] <- nrow(exprs(comb_batch))
j=2
for (i in 1:length(contrasts))
{
  stats_table[,j] <- length(which(diffexp[[i]]$adj.P.Val < 0.05))
  j=j+1
}
pander(stats_table,style="rmarkdown")
```

```{r message=FALSE, warning=FALSE}
# write output files
cont_names <- names(contrasts)
out_path <- '../../../data/processed/output/norm_with_acetylome_and_batch_correction_09_02_2022'
FP_out_path <- '../../../data/processed/output/FP_batch_effect_removed/'
if(!dir.exists(out_path)){
  dir.create(out_path, recursive = T)
}
for( x in 1:length(cont_names) ){
  each_cont <- cont_names[x]
  de_tab <- diffexp[[x]]
  out_file <- paste(each_cont, 'v1.0.csv', sep='_')
  out_file_p <- paste( out_path, out_file, sep='/')
  
  # add full proteome results to acetylome results
  FP_de_file <- str_c(FP_out_path,out_file, sep = '/' )
  FP_tab <- read_csv(FP_de_file) %>% 
    dplyr::select(Accessions, FP_log2FC=log2FC, FP_adj.P.Val=adj.P.Val)
  
  comb_tab <- inner_join(de_tab,FP_tab, by='Accessions')
  
  write_csv(x=comb_tab, file = out_file_p)
  
  hist(comb_tab$P.Value, main=each_cont, xlab='p-values')
}

```



## Software version.

```{r software_version, warning=FALSE, message=FALSE}
packageVersion("qPLEXanalyzer")
```